---
title: "Prediction Assignment Writeup"
author: "Anirban Chakraborty"
date: "3 January 2018"
output: 
    prettydoc::html_pretty:
        theme: cayman
        highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, message=FALSE, results = 'asis')
```

# **Introduction**
Using devices such as _Jawbone Up_, _Nike FuelBand_, and _Fitbit_ it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data
The training data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv. 
The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

## Goal
The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. I will create this report describing how the model has been built, how cross validation has been used, what is the expected out of sample error is, and why I made the choices I did.

```{r loading_libraries, echo=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(RColorBrewer)
library(randomForest)
library(knitr)
```

### Getting and Loading data
```{r loading_data}
set.seed(12345)
# download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "pml-training.csv")
# download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "pml-testing.csv")
training <- read.csv("pml-training.csv")
testing  <- read.csv("pml-testing.csv")
```

Partitioning the training dataset into training and cross-validation sets
```{r partition}
inTrain <- createDataPartition(training$classe, p = 0.6, list = FALSE)
myTraining <- training[inTrain,]
myTesting  <- training[-inTrain,]
dim(myTraining)
dim(myTesting)
```

### Cleaning the data
Removing NearZeroVariance variables
```{r remove_nzv}
nzv <- nearZeroVar(myTraining, saveMetrics = TRUE)
myTraining <- myTraining[,nzv$nzv == FALSE]

nzv <- nearZeroVar(myTesting, saveMetrics = TRUE)
myTesting <- myTesting[,nzv$nzv == FALSE]
```

Removing the first column of myTraining dataset
```{r remove_first_col}
myTraining <- myTraining[c(-1)]
```

Cleaning variables with more than 60% NA
```{r cleaning_var}
trainingV3 <- myTraining
for(i in 1:length(myTraining)) {
    if(sum(is.na(myTraining[,i]))/ nrow(myTraining) >= 0.7) {
        for(j in 1:length(trainingV3)) {
            if(length(grep(names(myTraining[i]), names(trainingV3)[j])) == 1) {
                trainingV3 <- trainingV3[, -j]
            }
        }
    }
}
myTraining <- trainingV3
```

Transform myTraining and testing datasets
```{r transform}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58])   #removing classe column
myTesting <- myTesting[clean1]
testing <- testing[clean2]
dim(myTesting)
dim(testing)
```

Coerce the data into same type
```{r coerce}
for (i in 1:length(testing)) {
    for (j in 1:length(myTraining)) {
        if(length(grep(names(myTraining[i]), names(testing)[j])) == 1) {
            class(testing[j]) <- class(myTraining[i])
        }
    }
}

testing <- rbind(myTraining[2, -58], testing)
testing <- testing[-1, ]
```

### Prediction with Decision Trees
```{r prediction, fig.width= 20, fig.height= 15}
set.seed(12345)
modFitA1 <- rpart(classe ~ ., data = myTraining, method = "class")
fancyRpartPlot(modFitA1)
```

```{r cmtree}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
cmtree <- confusionMatrix(predictionsA1, myTesting$classe)
cmtree
```

```{r decision_tree}
plot(cmtree$table, col = cmtree$byClass, main = paste("Decision Tree Confusion Matrix: Accuracy =", round(cmtree$overall['Accuracy'], 4)))
```

### Prediction with Random Forest
```{r random_forest}
set.seed(12345)
modFitB1 <- randomForest(classe ~ ., data = myTraining)
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
cmrf <- confusionMatrix(predictionsB1, myTesting$classe)
cmrf
```

```{r plot_rf}
plot(modFitB1)
plot(cmrf$table, col = cmtree$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmrf$overall['Accuracy'], 4)))
```

### Prediction with Generalized Boosted Regression
```{r pred_gbm}
set.seed(12345)
fitControl <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
gbmFit1 <- train(classe ~., data = myTraining, method = "gbm", trControl = fitControl, verbose = FALSE)

gbmPredTest <- predict(gbmFit1, newdata = myTesting)
gbmAccuracyTest <- confusionMatrix(gbmPredTest, myTesting$classe)
gbmAccuracyTest
```

```{r plot_gbm}
plot(gbmFit1, ylim = c(0.9,1))
```

### Prediction Results on the Test data
Random Forests gave an Accuracy in the myTesting dataset of 99.89%, which was more accurate that what I got from the Decision Trees or GBM. The expected out-of-sample error is 100-99.89 = 0.11%.

```{r final_predict}
predictionsB2 <- predict(modFitB1, testing, type = "class")
predictionsB2
```











